Automatic Parallelization and Transparent Fault Tolerance

### 2016-07-14 pubs/tfp16-draft/paper.tex
paper minor edits

An almost unbounded amount of interesting development suggests itself; one
such path could in principle lead to the recreation of GHC\@.  However, given
the minuscule budget (and therefore the employment of student interns) we must
maintain sharp focus on the main goals: implementing and evaluating mechanisms
for automatic parallelism and fault tolerance.  Currently we have
a fully functioning serial implementation including
\begin{itemize}
\item STG parser,
\item mini-Haskell parser, mini-Haskell to STG transformer,
\item type inferencer,
\item various analysis and transformation passes,
\item code generator,
\item runtime system including a garbage collector,
\item continuous integration and testing system,
\item extensive test suite, automated testing.
\end{itemize}

The \emph{mini-Haskell} front-end is detailed in Appendix~\ref{app:minihaskell}.

@@

For simplicity the implementation is 64-bit only.
%
The current, primitive parallel runtime is based on Pthreads and a
non-blocking work queue~\cite{Michael:1996}.  We are currently developing a
more realistic parallel runtime, experimenting with using the Argobots
user-level thread library~\cite{Seo:2015}.
%
Bridging the gap between GHC STG and our STG, and the corresponding
augmentation of the runtime is showing early signs of success.
%
An instrumentation infrastructure, integrated with the testing framework to
record and present time and space run-time statistics such as number of heap
and stack allocations and maximum stack depth, is planned for the near future.
%
The fault-tolerance mechanisms remain future work.
%
Parallel garbage collection is likely out of scope.
%
%We have implemented a simple copying garbage collector~\cite{Cheney:1970} with
%small \emph{nurseries} in the multithreaded case~\cite{Marlow:2008}. Marlow:2011 as well
%\cite{Harris:2005,Marlow:2008,Marlow:2009}

@@

\subsubsection{Strict but still pure.}
\vspace{-0.1in}
%\paragraph{Strict but still pure.}
The concept of strict but still pure in some sense dates back to the concept
of applicative order reduction in the lambda calculus.  In terms of a
notionally concrete programming language the first thorough explications may
be Backus' FP~\cite{Backus:1978} and follow-on FL
implementation~\cite{AikenFL,FLreport89}.
%
A notable argument for the potential merits of a strict-by-default
pure functional language over a non-strict one is by Sheard, who also provided
a very simple embedded implementation~\cite{Sheard:2003}.
%
A plug-in for GHC (prior to 8.0) exists that makes function bodies strict in
their non-recursive \texttt{let} bindings using a transformation on GHC Core
that recursively expands those \texttt{let} bindings into \texttt{case}
bindings~\cite{Bolingbroke:2008,strict-ghc-plugin}.
%
The Disciplined Disciple Compiler project appears to intend implementation of
a strict, higher order, functional language (or family of languages) compiler
and has been underway for some time, but the emphasis appears to be on a much
more complex type system for handling stateful computation, and the prospects
for its eventual completion are unknown~\cite{disciplined-disciple}.

@@

\section{(Ab)Using GHC}

\subsection{Producing Full STG Programs}

When compiling a standard Haskell program with GHC, modules imported
from other packages have already been compiled, and GHC simply links
the user object code with them.  Our compiler has no facility for
linking GHC-produced object code, so we need it to produce
\textit{full} STG programs built up from primitive operations.  GHC
does not offer object-to-STG translation, so this means we must
produce our own Prelude and supporting Base libraries to be included
with every program, and decoupling Haskell from GHC's wired-in types
as much as possible.

This is partially accomplished with extensions like
\texttt{NoImplicitPrelude} and \texttt{RebindableSyntax} (which
implies the former).  We cannot get a complete separation, however:
Numeric literals are tied to GHC's \texttt{Integer} and
\texttt{Rational} types, and guards are tied to GHC's \texttt{Bool}.
Lists and tuples, being special syntactic forms are also tightly
wired-in. To get around this, we provide placeholder definitions that
parallel any that we cannot replace, and use them to crudely translate
the exceptional cases.

\subsection{Navigating GHC's Internals}

One of the goals of using GHC as a library was to avoid being
overwhelmed with the necessary complexity of such a large and mature
tool, but this has not been entirely unavoidable.  Simply producing
STG without incurring any (non-temporary) file creation proved not to
be a difficult task, and can be accomplished using only the interface
exposed via the primary API, the \texttt{GHC} module.  This abstract
STG is even ostensibly very close to our own, but the auxiliary data
structures (e.g. those encapsulating identifier information and types)
are rich and deeply nested.  Determining what is strictly necessary
for our backend and what we have no implementation of (e.g. the
Foreign Function Interface) has taken considerable time and effort.

@@

>>>>

An almost unbounded amount of interesting development suggests itself; one
such path could in principle lead to the recreation of GHC\@.  However, given
the time constraints we must maintain sharp focus on the main goals:
implementing and evaluating mechanisms for automatic parallelism and fault
tolerance.  Currently we have a fully functioning serial implementation
including
\begin{itemize}
\item STG parser,
\item mini-Haskell parser, mini-Haskell to STG AST transformer,
\item partial coupling with the GHC front-end,
\item type inferencer,
\item various analysis and transformation passes,
\item code generator,
\item runtime system including a (serial) garbage collector and instrumentation framework,
\item continuous integration and testing system,
\item extensive test suite, automated testing.
\end{itemize}
Appendixes~\ref{app:minihaskell} and~\ref{app:ghclibrary}
describe the \emph{mini-Haskell} front-end and our approach to using GHC as a front end, respectively.
%

While wall-clock time can be used to simply compare the relative efficiency of
execution under various evaluation strategies, we seek more detailed
information.  We have in place a simple infrastructure for optionally logging
a thorough collection of run-time statistics such as number, size, and type of heap
and stack allocations, as well as various garbage collection statistics.  We also
measure cumulative time spent in various operations.  Future work includes processing
these measurement in an other than ad hoc fashion.

@@

For simplicity the implementation is 64-bit only.
%
The current, primitive parallel runtime is based on Pthreads and a
non-blocking work queue~\cite{Michael:1996}.  We are currently developing a
more realistic parallel runtime, experimenting with using the Argobots
user-level thread library~\cite{Seo:2015}.
%
%Bridging the gap between GHC STG and our STG, and the corresponding
%augmentation of the runtime is showing early signs of success.
%
An instrumentation infrastructure, integrated with the testing framework to
record and present time and space run-time statistics such as number of heap
and stack allocations and maximum stack depth, is planned for the near future.
%
The fault-tolerance mechanisms remain future work.
%
Parallel garbage collection is likely out of scope.
%
%We have implemented a simple copying garbage collector~\cite{Cheney:1970} with
%small \emph{nurseries} in the multithreaded case~\cite{Marlow:2008}. Marlow:2011 as well
%\cite{Harris:2005,Marlow:2008,Marlow:2009}

@@

\subsubsection{Strict but still pure.}
\vspace{-0.1in}
%\paragraph{Strict but still pure.}
The concept of strict but still pure in some sense dates back to the concept
of applicative order reduction in the lambda calculus.  In terms of a
notionally concrete programming language the first thorough explications may
be Backus' FP~\cite{Backus:1978} (building on Landin's ``theory''~\cite{Landin64}).
% and follow-on FL implementation~\cite{AikenFL,FLreport89}.
%
A notable argument for the potential merits of a strict-by-default
pure functional language over a non-strict one is by Sheard, who also provided
a very simple embedded implementation~\cite{Sheard:2003}.
%
A plug-in for GHC (prior to 8.0) exists that makes function bodies strict in
their non-recursive \texttt{let} bindings using a transformation on GHC Core
that recursively expands those \texttt{let} bindings into \texttt{case}
bindings~\cite{Bolingbroke:2008,strict-ghc-plugin}.
%
The Disciplined Disciple Compiler project appears to intend implementation of
a strict, higher order, functional language (or family of languages) compiler
and has been underway for some time, but the emphasis appears to be on a much
more complex type system for handling stateful computation, and the prospects
for its eventual completion are unknown~\cite{disciplined-disciple}.

@@

@@

\section{Using GHC as a Library}
\label{app:ghclibrary}

%\subsection{Producing Full STG Programs}
\subsubsection{Producing full STG programs.}
When compiling a standard Haskell program with GHC, modules imported
from other packages have already been compiled, and GHC simply links
the user object code with them.  Our compiler has no facility for
linking GHC-produced object code, so we need it to produce
\textit{full}\/ STG programs built up from primitive operations.  GHC
does not offer object-to-STG translation, so this means we must
produce our own Prelude and supporting Base libraries to be included
with every program, and decoupling Haskell from GHC's wired-in types
as much as possible.

This is partially accomplished with extensions like
\texttt{NoImplicitPrelude} and \texttt{RebindableSyntax} (which
implies the former).  We cannot get a complete separation, however:
numeric literals are tied to GHC's \texttt{Integer} and
\texttt{Rational} types, and guards are tied to GHC's \texttt{Bool}.
Lists and tuples, being special syntactic forms, are also tightly
wired-in. To get around this we provide placeholder definitions that
parallel any that we cannot replace, and use them to crudely translate
the exceptional cases.

%\subsection{Navigating GHC's Internals}
\subsubsection{Navigating GHC's internals.}
One of the goals of using GHC as a library was to avoid
the internal complexity of such a large and mature
tool, but this has not been entirely unavoidable.  Simply producing
STG without incurring any (non-temporary) file creation proved not to
be a difficult task, and can be accomplished using only the interface
exposed via the primary API, the \texttt{GHC} module.  This abstract
STG is even ostensibly very close to our own, but the auxiliary data
structures (e.g. those encapsulating identifier information and types)
are rich and deeply nested.  Determining what is strictly necessary
for our backend and what we have no implementation of (e.g. the
Foreign Function Interface) has taken considerable time and effort.
